---
title: ess 0.1.1 is out!
author: Jorge Cimentada
date: '2018-03-04'
slug: ess-0-1-1-is-out
tags: ['R', 'packages']
---

```{r, echo = FALSE, message = FALSE, error = FALSE, warning = FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(tidyverse)
library(ess)
```

The new version of the ess package is out! `ess 0.1.1` fixes some bugs and inconsistencies across the package and has one important new feature and a change that breaks backward compatibility. You can see all changes [here](https://cimentadaj.github.io/ess/news/index.html).

Install the latest version `0.1.1` from CRAN with `install.packages("ess")`.

## New features

When downloading any round(s) from the European Social Survey the files are always accompanied by a script that recodes values like 6, 7, 8 and 9 or 96, 97, 98 and 99 to missings, depending on the question. This is a bit tricky because a question with a scale from 1 to 5 will have 6 to 9 as missing values and a question with a scale from 1 to 10 will have the missing values set as 96 to 99. The new `remove_missings()` function removes all missing values from all questions.

For example...

```{r}
library(tidyverse)
library(ess)

clean_df <-
  ess_rounds(1, "your_email@gmail.com") %>%
  recode_missings()
```
... will set all the missing categories to NA. That is, the 6 to 9 and 96 to 99 categories on the specific questions. It gives you flexibility in recoding specific categories such as 'Don't Know', 'Refusal' or both.

```{r}
another_clean_df <-
  ess_rounds(1, "your_email@gmail.com") %>%
  recode_missings(c("Refusal", "No answer"))
```

See `?recode_missings` for the missing categories that are available for recode.

However, I do not advise recoding missing values right away if you're exploring the dataset. If you want to manually recode missing values you can use the `recode_numeric_missing()` and `recode_strings_missing` correspondingly on numeric and string variables. They work the same as `recode_missings` but accept a vector of class labelled, the class of each of the columns that returns the `ess_*` functions.

For example

```{r}
another_clean_df$tvtot <-
  recode_numeric_missing(
    another_clean_df$tvtot,
    "Don't know"
    )
```

works for recoding the "Don't know" category. By default all missing values are chosen.

Note that both sets of functions **only** work with labelled numeric vectors from the `haven` package. If you use the `ess` package that's taken care of. If you download the data manually, you must read it with the `haven` package for these functions to work.

There are also two new `show_*` functions, namely `show_themes` and `show_rounds_country`.

The first one returns all available themes...
```{r, eval = TRUE}
show_themes()
```
... but doesn't haven a corresponding `ess_*` function. This means that it works purely for descriptive purposes.

Additionaly, `show_rounds_country` returns all countries that participated in a give round.

```{r, eval = TRUE}
show_rounds_country(rounds = 2)
```

You could use this to see which countries participated in all rounds. For example..
```{r, eval = TRUE}
all_countries <-
  map(show_rounds(), ~ show_rounds_country(.x)) %>%
  reduce(intersect)

all_countries
```

## Breaking changes

Finally, there is one change that breaks backward compatability. All the `ess_*` functions always used to return a list, regardless of the number of rounds that were requested. Now, `ess_*` functions return a `tibble` whenever it is request only one round and a list when more than one round is requested.

For example

```{r}
ess_rounds(1, "your_email@gmail.com")
```

will return a tibble but...

```{r}
ess_rounds(1:3, "your_email@gmail.com")
```

...will return a list with each tibble in a slot.

For more concrete examples check out the new website of the ess [here](https://cimentadaj.github.io/ess/). If you have any ideas for features or find a bug, please report [here](https://github.com/cimentadaj/ess/issues).