---
title: Brief Analysis of Independent/Unionist Vote in Catalonia
author: Jorge Cimentada
date: '2017-12-14'
slug: brief-analysis-independence-vote-catalonia
categories: []
tags: ['R', 'EDA']
---
```{r, echo = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

On a train from Barcelona-Madrid I started working with an `R` package called `ggrides`. To my surprise, the package contains one dataset that documents the change in independent/unionist vote for all Catalan municipalities from 1980 to 2015. This is very cool! Needless to say, I left what I was doing and started to dive into the dataset.

The data looks like this:

```{r}
library(ggridges)
Catalan_elections
```

Very straight forward. It's the 'Indy' or Independent vote and the 'Unionist' vote from 1980 until 2015. The data is complete for nearly all Municipalities, meaning that the data is available for all years. Only a handful (~ 40) do not have data starting from 1980.

Basically, I had two ideas. First, how have municipalities changed over time? Has the independence vote grown over time? These two questions are very interesting considering that the topic is very hot. In fact, next week new Catalan elections will be held in a scenario never seen before. The elections are predicted to be very contested with Independent parties losing some votes.

Well, first

```{r}
library(scales)
library(tidyverse)


Catalan_elections$Option <- with(Catalan_elections, ifelse(Option == "Indy", "Independencia", "Union"))

# Which municipalities have decreased/increased indy vote over time
# tile map

avg_pl <-
  Catalan_elections %>%
  group_by(Municipality, Option) %>%
  summarize(first_year = first(Percent, Year),
            last_year = last(Percent, Year)) %>%
  group_by(Option) %>%
  summarize(first_year = median(first_year, na.rm = TRUE),
            last_year = median(last_year, na.rm = TRUE)) %>%
  mutate(id = 1:nrow(.)) %>%
  gather(year, value, -id, -Option)

bcn_pl <-
  Catalan_elections %>%
  filter(Municipality == "Barcelona") %>%
  group_by(Municipality, Option) %>%
  summarize(first_year = first(Percent, Year),
            last_year = last(Percent, Year)) %>%
  mutate(id = 1:nrow(.)) %>%
  gather(year, value, ends_with("year"))

base_plot <-
  Catalan_elections %>%
  group_by(Municipality, Option) %>%
  summarize(first_year = first(Percent, Year),
            last_year = last(Percent, Year)) %>%
  mutate(id = paste0(Municipality, "_", Option)) %>%
  gather(year, value, ends_with("year")) %>%
  ggplot(aes(year, value)) +
  geom_point(alpha = 0.1, size = 2) +
  geom_line(aes(group = id), alpha = 0.1)


median_plot <-
  base_plot +
  geom_point(data = avg_pl, aes(year, value),
            colour = "red", alpha = 0.5, size = 2) +
  geom_line(data = avg_pl, aes(year, value, group = id),
            colour = "red", alpha = 0.5, size = 2)

bcn_plot <-
  median_plot +
  geom_point(data = bcn_pl, aes(year, value),
             colour = "blue", alpha = 0.5, size = 2) +
  geom_line(data = bcn_pl, aes(year, value, group = id),
            colour = "blue", alpha = 0.5, size = 2)



pretty_plot <-
  bcn_plot +
  scale_x_discrete(name = NULL,
                   labels = c("1980",
                              "2015")) +
  scale_y_continuous(name = "% de votos a favor de:",
                     breaks = seq(0, 100, 20),
                     labels = percent(seq(0, 1, 0.2))) +
  facet_wrap(~ Option, strip.position = "bottom") +
  labs(
    title = "Aumento de voto independentista/unionista en tres decadas",
    subtitle = "Linea roja es la mediana de todos los municipios - Linea azul es Barcelona",
    caption = "Datos obtenidos por @marcbeldata - Gr치fico y an치lisis por @cimentadaj"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, family = "Arial-BoldMT"),
    plot.subtitle = element_text(size = 16, color = "#666666"),
    plot.caption = element_text(size = 12, color = "#666666"),
    strip.text.x = element_text(size = 14),
    axis.title.y = element_text(size = 16),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 14)
  )

pretty_plot

# There are 16 times more municipalities that have always been independent
# than municipalities that have always been unionist

```


```{r}
# More specifically
Catalan_elections %>%
  filter(Municipality == "Barcelona") %>%
  ggplot(aes(Year, Percent, group = Option, colour = Option)) +
  geom_line(alpha = 0.5, size = 2) +
  scale_x_continuous(name = NULL) +
  scale_colour_discrete(name = NULL) +
  scale_y_continuous(name = "% de votos a favor de:",
                     lim = c(0, 100),
                     breaks = seq(0, 100, 20),
                     labels = percent(seq(0, 1, 0.2))) +
  labs(
    title = "Aumento de voto independentista/unionista en Barcelona",
    caption = "Datos obtenidos por @marcbeldata - Gr치fico y an치lisis por @cimentadaj"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, family = "Arial-BoldMT"),
    plot.subtitle = element_text(size = 16, color = "#666666"),
    plot.caption = element_text(size = 12, color = "#666666"),
    legend.position = "top",
    legend.text = element_text(size = 14),
    strip.text.x = element_text(size = 14),
    axis.title.y = element_text(size = 16),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 14)
  )
```