---
title: "ess is now essurvey"
author: "Jorge Cimentada"
date: '2018-03-26'
slug: ess-is-now-essurvey
tags: ['R', 'packages', 'project']
comments: no
showcomments: yes
showpagemeta: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


My `ess` package has be renamed to `essurvey`. For quite some time I've been pondering whether I should change the name. All of this comes from a dicussion we had on the [R-pkg mailing list](http://r.789695.n4.nabble.com/R-pkgs-Release-of-ess-0-0-1-td4746540.html) where many R users suggested that the name was unfortunate given that Emacs Speaks Statistics (ESS) has a long precedence in the R community and the names are very similar. Later on, when submitting the package to [rOpensci](https://ropensci.org/), Jim Hester [raised the fact once again](https://github.com/ropensci/onboarding/issues/201#issuecomment-372304003), without being aware of the previous email thread.

Considering that I was already changing some of the functionalities of the package due to the [rOpensci review](https://github.com/ropensci/onboarding/issues/201), I decided to change the package name and republish an improved version of `ess` as `essurvey 1.0.0`. `essurvey` is now on CRAN and the repository has been moved to rOpensci's [github account](https://github.com/ropensci/essurvey).

The new package is mostly similar although there are now some deprecated functions and new features. Below are the main changes.

* You can login **once** using `set_email("your_email")` and avoid rewriting your email in every call to the `ess_*` functions.

* All `ess_*` functions have been deprecated in favour of similar `import_*` functions. For example:

```{r}
ess_rounds(1:7)
```

becomes..

```{r}
import_rounds(1:7)
```

But that's the same you would say. The only difference is that with `ess_rounds` you could download data in Stata, SPSS or SAS formats directly. For that, there's now the `download_*` functions.

```{r}
download_rounds(
  1:5,
  output_dir = getwd(),
  format = "spss"
)
```

All of the above applies to `ess_country` and the `ess_all_*` functions. There's also some other minor changes you can checkout in the [NEWS](https://github.com/ropensci/essurvey/blob/master/NEWS.md) file. If you haven't tried `essurvey`, you can visit the package website for more detailed examples at https://ropensci.github.io/essurvey/.
