---
title: Locating parts of a string with `stringr`
author: Jorge Cimentada
date: '2019-12-08'
slug: locating-parts-of-a-string-with-stringr
categories: ["R"]
tags: ["R", "regex"]
comments: no
showcomments: yes
showpagemeta: yes
---

I was wondering the realms of StackOver Flow answering some questions when I encoutered a question that looked to extract some parts of a string based on a regex. I thought I knew how to do this with the package `stringr` using, for example,  `str_sub` but I found it a bit difficult to map how `str_locate` complements `str_sub`. 

`str_locate` and `str_locate_all` give back the locations of your regex inside the desired string as a `matrix` or a `list` respectively. However, that didn't look very intuitive to pass on to `str_sub` which (I thought) only accepted numeric vectors with the indices of the parts of the strings that you want to extract. However, to my surprise, `str_sub` accepts not only numeric vectors but also a matrix with two columns, precisely the result of `str_locate`. 

Let's create a set of random strings from which we want to extract the word `special*word`, where `*` represents a random number.

```{r }
library(stringr)    

test_string <-
  replicate(
    100,
    paste0(
      sample(c(letters, LETTERS, paste0("special", sample(1:10, 1),"word")), 15),
      collapse = "")
  )

head(test_string)

```

Using `str_locate` returns a matrix with the positions of all matches for **every string**. This is what's called **vectorised** functions in R. 

```{r }
location_matrix <-
  str_locate(test_string, pattern =  "special[0-9]word")

head(location_matrix)

```

For this example this wouldn't work, but I was also interested in checking how the result of `str_locate_all` would fit in this workflow. `str_locate_all` is the same as `str_locate` but since it can find **more** than one match per string, it returns a list with the same slots as there are strings in `test_string` with a matrix per slot showing the indices of the matches. Since many of the strings in `test_string` might not have `special*word`, we need to fill out those matches with `NA`:

```{r}
location_list <-
  str_locate_all(test_string, pattern =  "special[0-9]word") %>% 
  lapply(function(.x) if (all(is.na(.x))) matrix(c(NA, NA), ncol = 2) else .x) %>%
  {do.call(rbind, .)}

head(location_list)

```

Now that we have everything ready, `str_sub` can give our desires results using both numeric vectors as well as the entire matrix:

```{r}
# Using numeric vectors from str_locate
str_sub(test_string, location_matrix[, 1], location_matrix[, 2])
# Using numeric vectors from str_locate_all
str_sub(test_string, location_list[, 1], location_list[, 2])
# Using the entire matrix
str_sub(test_string, location_matrix)
```

A much easier approach to doing the above (which is cumbersome and verbose) is to use `str_extract`:

```{r }
str_extract(test_string, "special[0-9]word")
```

However, the whole objecive behind this exercise was to clearly map out how to connect `str_locate` to `str_sub` and it's much clearer if you can pass the entire matrix. However, converting `str_locate_all` is still a bit tricky. 
